Basics of cubic spline models

### 2021-11-25 text/21/cubic-splines.Rmd
More tinkering with cubic-splines.Rmd

```{r}
display_terms <- function(df, lb) {
  names(df) <- c("x", "y")
  integer_data <- df[1:21, ]
  ggplot(df, aes(x, y)) +
      ggtitle(lb) +
      geom_line() +
      geom_point(data=integer_data, aes(x, y)) +
      xlab(" ") +
      ylab(" ")
}

@@

for (j in 1:16) {
  xm %>% 
    select(c(17, j)) %>%
    display_terms(lb[j]) -> cplot
  plot(cplot)
}

@@

```{r}
display_fit <- function(model_fit, lb) {
  model_fit %>%
    augment(newdata=xm) %>%
      ggplot(aes(c2, .fitted)) +
        ggtitle(lb) +
        geom_line() +
        geom_point(aes(c2, y), na.rm=TRUE) +
      xlab(" ") +
      ylab(" ")
}

lm(y~c2+c3+c4, data=xm) %>%
  display_fit("Single cubic fit")
```

@@

lm(y~c2+c3+c4+c5+c6+c7+c8+c9+c10+c11+c12+c13+c14+c15+c16, data=xm)  %>%
  display_fit("Discontinous cubic splines")
```

@@

```{r}
kable(xm[1:20, c(1:4, 6:8, 10:12, 14:16)])
lm(y~c2+c3+c4+c6+c7+c8+c10+c11+c12+c14+c15+c16, data=xm)  %>%
  display_fit("Continous but not smooth cubic splines")
```

Smoothness is measured in terms of the continuity of derivatives. Here is a function that has a continuous first derivative.

```{r}
kable(xm[1:20, c(1:4, 7:8, 11:12, 15:16)])
lm(y~c2+c3+c4+c7+c8+c11+c12+c15+c16, data=xm) %>%
  display_fit("Continous and Smooth (1st derivative) splines")

@@

```{r}
kable(xm[1:20, c(1:4, 8, 12, 16)])
lm(y~c2+c3+c4+c8+c12+c16, data=xm) %>%
  display_fit("Continous and smooth (1st and 2nd derivatives) cubic splines")
```

While this approach is simple, there is an issue with multicollinearity.

@@

This approach is simple and easy to follow, but there is one catch. The correlations are quite high and this can lead to computational problems, including rounding errors. So most spline models implemented on a computer use a different approach.

B-splines provide a solution with less issues of multi-collinearity.

@@

```{r}
display_matrix <- function(df, lb) {
  names(df) <- c("x", "y")
  integer_data <- df[1:21, ]
  ggplot(df, aes(x, y)) +
      ggtitle(lb) +
      geom_line() +
      geom_point(data=integer_data, aes(x, y)) +
      xlab(" ") +
      ylab(" ")
}
for (j in 1:dim(linear_spline)[2]) {
  cbind(xm$c2, linear_spline[ , j]) %>% 
    data.frame %>%
    display_matrix("Linear spline terms") %>%
    plot
}
```

```{r}
lm(xm$y~linear_spline) %>%
  display_fit("Linear splines")
```

```{r}
quadratic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=2, intercept=TRUE)
kable(quadratic_spline[1:21, ])

for (j in 1:dim(quadratic_spline)[2]) {
  cbind(xm$c2, quadratic_spline[ , j]) %>% 
    data.frame %>%
    display_matrix("Quadratic spline terms") %>%
    plot
}
```

```{r}
lm(xm$y~quadratic_spline) %>%
  display_fit("Quadratic splines")
```

@@

```{r}
for (j in 1:dim(cubic_spline)[2]) {
  cbind(xm$c2, cubic_spline[ , j]) %>% 
    data.frame %>%
    display_matrix("Cubic spline terms") %>%
    plot
}
```

```{r}
lm(xm$y~cubic_spline) %>%
  display_fit("Cubic splines")
```

@@

```{r}
for (j in 1:dim(natural_spline)[2]) {
  cbind(xm$c2, natural_spline[ , j]) %>% 
    data.frame %>%
    display_matrix("Natural spline terms") %>%
    plot
}
```

@@

>>>>

```{r}
display_curve <- function(dat, lb) {
  # Plot a line connecting the points in dat
  # Assume that column 1 is x and column 2 is y
  # Work with either a data frame or matrix
  dat %>%
    data.frame %>%
    select(1:2) %>%
    set_names(c("x", "y")) %>%
    ggplot(aes(x, y)) +
      ggtitle(lb) +
      geom_line() +
      xlab(" ") +
      ylab(" ")
}

display_dots <- function(cplot, dat) {
  # Add dots in dat to an existing graph
  # Assume that column 1 is x and column 2 is y
  # Work with either a data frame or matrix
  dat %>%
    data.frame %>%
    set_names(c("x", "y")) -> df
  cplot + geom_point(data=df, aes(x, y))
}

@@

for (j in 1:16) {
  dat <- xm[ , c("x",  paste0("c", j))]
  display_curve(dat, lb[j]) %>%
    display_dots(dat[1:21, ]) %>%
    plot
}

@@

lm(y~c2+c3+c4, data=xm) %>%
  augment %>%
  select(c2, .fitted) %>%
    display_curve("Single cubic fit") %>%
    display_dots(simulated_example)
```

@@

```{r}
lm(y~c2+c3+c4+c5+c6+c7+c8+c9+c10+c11+c12+c13+c14+c15+c16, data=xm)  %>%
  augment %>%
  select(c2, .fitted) %>%
    display_curve("Discontinous cubic splines") %>%
    display_dots(simulated_example)
```

@@

### Continuous splines

```{r}
kable(xm[1:20, c(1:4, 6:8, 10:12, 14:16)])
lm(y~c2+c3+c4+c6+c7+c8+c10+c11+c12+c14+c15+c16, data=xm)  %>%
  augment %>%
  select(c2, .fitted) %>%
    display_curve("Continuous but not smooth cubic splines") %>%
    display_dots(simulated_example)
```

While this graph is continuos, it still takes some abrupt term. The mathemSmoothness is measured in terms of the continuity of derivatives.

### Smooth splines

Here is a function that has a continuous first derivative.

```{r}
kable(xm[1:20, c(1:4, 7:8, 11:12, 15:16)])
lm(y~c2+c3+c4+c7+c8+c11+c12+c15+c16, data=xm) %>%
  augment %>%
  select(c2, .fitted) %>%
    display_curve("Continuous and smooth (1st derivative) cubic splines") %>%
    display_dots(simulated_example)
```

@@

```{r}
kable(xm[1:20, c(1:4, 8, 12, 16)])
lm(y~c2+c3+c4+c8+c12+c16, data=xm) %>%
  augment %>%
  select(c2, .fitted) %>%
    display_curve("Continous and smooth (1st and 2nd derivatives) cubic splines") %>%
    display_dots(simulated_example)
```

This approach is simple and easy to follow, but there is one catch. There is an issue with multicollinearity.

@@

The correlations are quite high and this can lead to computational problems, including rounding errors. So most spline models implemented on a computer use a different approach.

### B-splines

B-splines provide a solution with less issues of multi-collinearity. You use the bs function in the splines package to compute B-splines.

#### Linear B-splines

Although most applications use cubic B-splines, they are available in other forms. Here is the linear spline.

@@

The terms in the linear B-splines are mostly piecewise linear functions. Notice how they are zero outside certain ranges.

```{r}
for (j in 1:dim(linear_spline)[2]) {
  dat <- cbind(xm$c2, linear_spline[ , j])
  display_curve(dat, "Linear B-spline terms") %>%
    display_dots(dat[1:21, ]) %>%
    plot
}
```

```{r}
round(cor(linear_spline), 1))
```

```{r}
lm(xm$y~linear_spline) %>%
  augment %>%
  bind_cols(xm$c2) %>%
  select(c2, .fitted) %>%
    display_curve("Linear spline fit to the data") %>%
    display_dots(simulated_example)
```

```{r}
quadratic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=2, intercept=TRUE)
kable(quadratic_spline[1:21, ])
```

```{r}
for (j in 1:dim(quadratic_spline)[2]) {
  dat <- cbind(xm$c2, quadratic_spline[ , j])
  display_curve(dat, "Quadratic B-spline terms") %>%
    display_dots(dat[1:21, ]) %>%
    plot
}
```

```{r}
lm(xm$y~quadratic_spline) %>%
  bind_cols(xm$c2) %>%
  select(c2, .fitted) %>%
    display_curve("Quadratic spline fit to the data") %>%
    display_dots(simulated_example)
```

@@

```{r}
for (j in 1:dim(cubic_spline)[2]) {
  dat <- cbind(xm$c2, cubic_spline[ , j])
  display_curve(dat, "Cubic B-spline terms") %>%
    display_dots(dat[1:21, ]) %>%
    plot
}
```

```{r}
lm(xm$y~cubic_spline) %>%
  bind_cols(xm$c2) %>%
  select(c2, .fitted) %>%
    display_curve("Cubic spline fit to the data") %>%
    display_dots(simulated_example)
```

@@

```{r}
extended_x <- seq(-5, 25, by=1/64)
extended_spline <- predict(natural_spline, extended_x)
for (j in 1:dim(natural_spline)[2]) {
  cbind(extended_x, extended_spline[ , j]) %>% 
    data.frame %>%
    display_matrix("Natural spline terms") %>%
    plot
}
```

@@

References

Donald H. House. Chapter 14. Spline Curves. Available in [pdf format][hou1].

Aris Perperoglou, Willi Sauerbrei, Michal Abrahamowicz, Matthias Schmid. A review of spline function procedures in R. BMC Medical Research Methodology 19, 46 (2019). DOI: 10.1186/s12874-019-0666-3. Available in [html format][per1] or [pdf format][per2].

[hou1]: https://people.cs.clemson.edu/~dhouse/courses/405/notes/splines.pdf

[per1]: https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-019-0666-3
[per2]: https://bmcmedresmethodol.biomedcentral.com/track/pdf/10.1186/s12874-019-0666-3.pdf
